---
title: "R introduction"
author: "Mateusz Żółtak"
date: "11 stycznia 2016"
output: 
  html_document: 
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# What is R?

R can denote:

* programming language
    * the real name of the language is S (although R interpreter provides some extensions)
    * there are many S interpreters: S-PLUS, R, RevolutionR and others
* software for (mainly) statistical analysis
    * https://www.r-project.org/
    * originated as an open source interpreter of the S language
    * most features provided by community as a so called *packages*
* research environment
    * *R environment* (existing packages and IDEs) allow to conduct the whole research process in one application starting from data gathering, through analysis, ending up with writing papers
    
# R's pros and cons

Pros:

* for free
* almost all statistical methods you can imagine are available
* R's ecosystem provides a complete research environment, e.g.
    * powerful graphs capabilities
    * embeding R in markdown, LaTeX or on webpages
    * git/svn support, tests, etc for supporting reproducibility (and automation)
* great users community

Cons:

* it is a programming language you have to learn  
    * and the learning curve is said to be steep

# R syntax basics

We can divide language constructs into few categories:

* control structures  
  reserved grammar structures for steering program workflow
    * e.g. *if* stucture, loops, etc.
* operators  
  e.g. addition (_+_), substraction (_-_), multiplication (_*_), etc.
* variables  
  simply a piece of data with a name
* functions  
  "more sophisticated operators" used to generate and/or process data
    * from technical point of view functions in R are also variables (they are so called 1st class objects) but until you rich quite high skills in R it is far easier to think about them as something separate

## Operators

* Arithmetic: 
    * basic: _+, -, *, /_
    * power: _^_ 
        * *2^3* gives *8*
    * modulo division: _%%_
        * *57 %% 10* gives *7*
    * logical: _&, |, !_ (and, or, not)
* Comparison:
    * basic: _<, <=, ==, !=, >=, >_
    * inclusion: _a %in% b_ 
        * _1 %in% c(1, 2, 3)_ gives _TRUE_
* Many others we will skip at the moment

## Variables

There are 3 main value types:

* character  
    * character values are denoted by surrounding them with quotation marks (either ' or "")  
    * e.g. *'my strings'* or *"my string"*
* number (in general there is no distinction between integers and floating point ones)
    * e.g. *123* or *1.23*
* boolean  
    * *TRUE* or *T* and *FALSE* or *F*  
    * later we will always use the long syntax *TRUE* or *FALSE*

Each variable is:

* *a vector* (an ordered set of values) of a given type
    * even what you would call a single value in R will be a vector of length 1
    * *c(firstValue, secondValue[, ...])* syntax is used to create a vector
    * although for a single value you can simply put the value and R will automatically promote it to a vector (see examples)
* or *a list* of values (where each value can be of different type)
    * *list(firstValue, secondValue[, ...])* syntax is used to create a list

**To assign a variable a *=* or *<-* operators are used** (there is no difference between them).

**To display variable simply type its name.**

Examples:
```{r}
# vector of numbers
numbers = c(1, 2, 3)
numbers
# "single value" which in fact is also a vector (see below)
singleNumber = 3
singleNumber

# strings
strings = c('hello', 'world')
strings
singleString = 'world'
singleString

# booleans
bools = c(TRUE, FALSE)
bools
bool = TRUE
bool

# list
myList = list(numbers, TRUE, list(1, 'a'))
myList
```

*Operators perform operations on whole vectors*
```{r}
num1 = 1:3
num2 = 5:7
num1
num2
num1 * num2
```

### More on vectors

You can access given elements of a vector using their indexes.  
You can also pass a vector of indexes.
```{r}
numbers = c(1, 3, 5, 7)
numbers[2]
numbers[c(2, 4)]
```
And there is a useful shorthand syntax for creating range of numbers: *from:to*, e.g.:
```{r}
numbers[2:4]
```

Each element of a vector can have a name.  
Names can be assigned during vector creation or afterwords, using a *names()* function (which returns a vector of names):
```{r}
# assigning names during vector creation
# (2nd element will be unnamed)
namedVector = c(name1 = 1, 2, name3 = 3)
namedVector
names(namedVector)
names(namedVector)[2] = 'missingName'
namedVector
```

If a vector has names, you can also access its elements by names:
```{r}
namedVector = c(name1 = 1, name2 = 2, name3 = 3)
namedVector[c('name1', 'name3')]
```

### More on lists

You can access given elements of a list in a same way then a vector. **Returned object will be always a list.**  
Also the same rules apply to list's elements names.

```{r}
myNamedList = list(name1 = 1:2, 5:6, name3 = c('a', 'b'))
names(myNamedList)
names(myNamedList) = c('n1', 'n2', 'n3')
names(myNamedList)
myNamedList[c(1, 3)]
myNamedList['n2']
```

*There are two special syntaxes to fetch single element of a list as a vector:*
```{r}
myList = list(a = 1:2,  b = 3:4,  c = 5:6)
# with a [[ ]]
myList[[2]]
# with a $name
myList$b
```

### Matices

Matrix is *a vector divided into columns*.

You can create matrix using the *matrix()* function.  
It takes a vector of values which will be splitted into columns, the number of rows and the number of columns.
```{r}
myMatrix = matrix(1:12, nrow = 3, ncol = 4)
myMatrix
```

You can split vector into rows instead of columns if you want:
```{r}
myMatrix = matrix(1:12, nrow = 3, ncol = 4, byrow = TRUE)
myMatrix
```

Matrix allow you to address elements using row and column indexes:
```{r}
myMatrix = matrix(1:12, nrow = 3, ncol = 4)
myMatrix[1:2, 3:4]
myMatrix[1:2, ]
```

Instead of simply _names_ matrix has _colnames_ and _rownames_.
Matrix name can be also assigned during matrix creation or afterwords.
```{r}
myMatrix = matrix(1:12, nrow = 3, ncol = 4, dimnames = list(c('r1', 'r2', 'r3'), c('c1', 'c2', 'c3', 'c4')))
myMatrix
colnames(myMatrix) = c('col1', 'col2', 'col3', 'col4')
myMatrix
```

R is able to perform arithmetic operations on whole matrices:
```{r}
matrix1 = matrix(1:4, nrow = 2, ncol = 2)
matrix2 = matrix(5:8, nrow = 2, ncol = 2)
matrix1
matrix2
matrix1 + matrix2
```

But matrix can be also threated as a normal vector:
```{r}
myMatrix = matrix(1:12, nrow = 3, ncol = 4, dimnames = list(c('r1', 'r2', 'r3'), c('c1', 'c2', 'c3', 'c4')))
myMatrix
myMatrix[c(1, 5, 7)]
```

### Data frames

Data frame is like a matrix applied to a list.

*If all elements of a list have exactly the same length*, we can treat it as a table displaying each element of a list as a table column.  
You can convert a list to a data frame using _as.data.frame()_ function or use a _data.frame()_ shorthand:
```{r}
myList = list('a' = 1:3, b = c('a', 'b', 'c'))
myList
as.data.frame(myList)
data.frame('a' = 1:3, b = c('a', 'b', 'c'))
```

Similary to matrices, you can subset data frames using rows and columns indexes/names:
```{r}
myDataFrame = data.frame('a' = 1:3, b = c('a', 'b', 'c'), c = 5:7)
myDataFrame[2:3, c('a', 'c')]

# be aware that if you select only one column, you will get not a data frame but a vector
myDataFrame[2:3, c('a')]
```

And you can still access data frame's columns with the list's _$columnName_ , _[]_ and _[[]]_ operators:
```{r}
myDataFrame = data.frame('a' = 1:3, b = c('a', 'b', 'c'), c = 5:7)
myDataFrame$b
myDataFrame[2]
myDataFrame[[2]]
```

## Functions

There are to many to describe them briefly.

We will learn the ones important for us during workshop.

### Creating your own functions

You can create your own functions using syntax:
```{r}
myFunction = function(param1, param2){
  result = param1 * param2
  # any other code goes here...
  return(result)
}
myFunction(1:2, 3:4)
```

Of course your functions can take any number of parameters (also no one).

If you want your function to return nothing at all, use _return(NULL)_ at the end.  

## Control structures

### If / else

Examples should be self descriptive:
```{r}
if(1 == 2){
  print('1 == 2')
}else if(1 == 3){
  print('1 == 3')
}else{
  print('else')
}

if(1 == 1){
  print('1 == 1')
}
```

And if you are familiar with a shorthand for _if a then b else c_ avaiable in some languages (e.g. _a ? b : c_ in C/C++/C#/Java or _b if a else c_ in Python), you should welcome R's version:
```{r}
ifelse(TRUE, 1, 2)
ifelse(FALSE, 1, 2)

# of course it works on vectors!
ifelse(c(TRUE, FALSE), 1:2, 2:3)
```

### Loops

There are two main loop types in R:

* a *for(element in vector)* loop
* a *while(expression)* loop

Examples
```{r}
for(i in 1:10){
  print(i)
}

i = 1
while(i <= 10){
  print(i)
  i = i + 1
}
```
